\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-15 paper.qmd
Rewrite the motivation of our paper

High-dimensional (high-D) data is widespread in many fields including ecology and bioinformatics [@Guo2023], in part because of new data collection technologies [@Johnstone2009; @ayesha2020overview]. Working with high-dimensional data poses considerable challenges due to the difficulty in visualizing beyond two dimensions [@Jia2022]. High-dimensional data also presents difficulties for model fitting [@Johnstone2009], both computationally and interpretation, each of which benefits from visualization.

To create visual representations of high-dimensional data, it is common to apply dimension reduction techniques. Linear methods such as principal component analysis (PCA) [@Karl1901] have been used for many years, and remain popular. Non-linear methods such as multi-dimensional scaling (MDS) [@Torgerson1967] have also been routinely used. In the past decade, there has merged many new techniques non-linear dimension reduction (NLDR), such as t-distributed stochastic neighbor embedding (tSNE) [@Laurens2008], uniform manifold approximation and projection (UMAP) [@Leland2018], designed to capture the complex and non-linear relationships present within high-dimensional data [@Johnstone2009].

@@

@@

The outline of this paper is as follows. The @sec-background provides an detailed overview of dimension reduction methods, triangulation, and tours. Building upon this foundation, the @sec-methods delves into the proposed algorithm, **quollr**, and its implementation details. In @sec-prediction, discusses the effectiveness of the learned low-dimensional manifold in accurately representing the complex high-dimensional data. Following that, @sec-simpleex presents simple examples from simulations to illustrate the functionality of the algorithm. Subsequently, @sec-applications showcases real-world applications of **quollr** on different data sets, particularly in single-cell RNA-seq data. These applications reveal insights into the performance and trustworthiness of NLDR algorithms. We analyze the results to identify situations where NLDR techniques may lead to misleading interpretations. Finally, @sec-conclusions concludes by summarizing the findings and emphasizing the significance of the proposed approach in tackling the challenges of high-dimensional data visualization.

@@

As a result of hexagonal binning for high-D data, all the high-D data are clustered into hexagons. In this step, the bin centroids ($C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2}$) (see @fig-meth Step 2) are obtained [@Carr2013].

@@

In this step, the algorithm proceeds to triangulate the hexagonal bin centroids (see @fig-meth (Step 3)). Triangulation is a fundamental process in computational geometry and computer graphics that involves dividing a set of points in a given space into interconnected triangles [@article30]. One common algorithm used for triangulation is Delaunay triangulation [@article26, @article54], where points are connected in a way that maximizes the minimum angles of the resulting triangles, leading to a more regular and well-conditioned triangulation.

@@

## Discussion {#sec-discussion}

Our research introduces a comprehensive framework that leverages tours for interactive exploration of high-dimensional data coupled with a low-dimensional manifold, facilitated by the `quollr` R package. Regardless of the Non-Linear Dimension Reduction (NLDR) technique in use, our approach demonstrates effectiveness through simulation examples, particularly in the iterative removal of long edges for a smoother representation and capturing cluster variance.

@@

In conclusion, our framework presents a powerful tool for researchers and analysts in single-cell studies to assess their embeddings by visually inspecting them alongside the original data. By leveraging the advantages of tours and low-dimensional manifolds, our approach offers valuable insights into the data transformation process, empowering users to make informed decisions in analyzing high-dimensional data. Future work could enhance the method's robustness in the presence of noise and explore additional evaluation metrics for quantifying embedding accuracy.

>>>>

High-dimensional (high-D) data is prevalent across various fields, such as ecology and bioinformatics [@Guo2023], due to advancements in data collection technologies [@Johnstone2009; @ayesha2020overview]. However, visualization of high-D data introduces significant challenges, because the complexity of visualizing data beyond two dimensions [@Jia2022]. In recent years, interactive and dynamic graphics systems like **liminal** [@article21] —which employs interactive tools like brushing and linking [@article58]—and software tools such as **XGobi**, **GGobi** [@article60], **tourr**  [@article61], **detourr** [@article22], and **langevitour**  [@article09], involving dynamic methods like tours [@Asimov1985], have played a key role in visualizing high-D data (data-vis).

To create low-dimensional representations (typically in 2D) of high-D data, it is common to apply dimension reduction (DR) techniques. Approaches for DR involve linear methods such as principal component analysis (PCA) [@Karl1901], non-linear methods such as multi-dimensional scaling (MDS) [@Torgerson1967]. In the past decade, many new non-linear dimension reduction (NLDR) techniques have emerged, such as t-distributed stochastic neighbor embedding (tSNE) [@Laurens2008] and uniform manifold approximation and projection (UMAP) [@Leland2018], designed to capture the complex and non-linear relationships present within high-D data [@Johnstone2009]. NLDR techniques are the 2D models of high-D data in our context.

Visualization of NLDR techniques (m-vis) [@article59] for the same high-D data is particularly important in understanding and finding the best representation. If we have done so, the 2D models can be quite different from the data structure shown in high-D space. Therefore, visualizing the 2D model in high-D space (m-in-dis) is more useful to answer different types of questions:

- Is there a best representation of high-D data or are they all providing  equivalent information? Is there a best parameter choice to fit the 2D model? How does the model change when it's parameters change?

- How well the does the 2D models capture the data structure? Is the model fitting able to capture different data structure like non-linear, clustering?

If we can't easily ask and answer these questions, the ability of understanding the models are restricted. To find the best 2D model and the parameter choices, better understanding of the underlying science is really important.

Also, the importance of m-vis along with data-vis has been recognized and incorporated into interactive software, **liminal** [@article21]. But the 2D model and high-D visualize side by side and interactive like brushing and linking connect the data in the two panels. To address this challenge, we propose a novel approach by combining the tour technique with a low-dimensional manifold. This manifold is created through the synergistic use of NLDR techniques, hexagonal binning, and triangulation. This integration facilitates a more understanding of the data structure, how well (or how poorly) NLDR techniques perform.

The outline of this paper is as follows. The @sec-background provides an detailed overview of dimension reduction methods, and tours. Building upon this foundation, the @sec-methods delves into the proposed algorithm, its implementation details, how to tune the model, model summaries, and a synthetic example to illustrate the functionality of the algorithm. Subsequently, @sec-applications showcases applications of the algorithm on different data sets, particularly in single-cell RNA-seq data. These applications reveal insights into the performance and trustworthiness of NLDR algorithms. We analyze the results to identify situations where NLDR techniques may lead to misleading interpretations. Finally, @sec-conclusions concludes by summarizing the findings and emphasizing the significance of the proposed approach in tackling the challenges of high-dimensional data visualization.

<!--High-dimensional (high-D) data is widespread in many fields including ecology and bioinformatics [@Guo2023], in part because of new data collection technologies [@Johnstone2009; @ayesha2020overview]. Working with high-dimensional data poses considerable challenges due to the difficulty in visualizing beyond two dimensions [@Jia2022]. High-dimensional data also presents difficulties for model fitting [@Johnstone2009], both computationally and interpretation, each of which benefits from visualization.

To create visual representations of high-dimensional data, it is common to apply dimension reduction techniques.

@@

The earlier approaches for DR involve linear methods such as PCA [@Karl1901]. PCA aims to maintain the second-order statistics of the data by projecting the points into the low dimensional space that preserves the maximum amount of variance among all such projections. As a result, PCA has been shown to be effective in preserving the global structure of the data (Silva & Tenenbaum, 2003).
The global structure includes the overall shape of the dataset, placement of the clusters, and the existence of potential outliers. Unlike PCA, much of the focus of the more recent non-linear methods, including t-SNE (Maaten & Hinton,2008), LargeVis (Tang et al., 2016), and UMAP (McInnes et al., 2018) has been on preserving the local neighborhood structure of each individual point. Mainly,

@@

The outline of this paper is as follows. The @sec-background provides an detailed overview of dimension reduction methods, triangulation, and tours. Building upon this foundation, the @sec-methods delves into the proposed algorithm, **quollr**, and its implementation details. In @sec-prediction, discusses the effectiveness of the learned low-dimensional manifold in accurately representing the complex high-dimensional data. Following that, @sec-simpleex presents simple examples from simulations to illustrate the functionality of the algorithm. Subsequently, @sec-applications showcases real-world applications of **quollr** on different data sets, particularly in single-cell RNA-seq data. These applications reveal insights into the performance and trustworthiness of NLDR algorithms. We analyze the results to identify situations where NLDR techniques may lead to misleading interpretations. Finally, @sec-conclusions concludes by summarizing the findings and emphasizing the significance of the proposed approach in tackling the challenges of high-dimensional data visualization. -->

@@

As a result of hexagonal binning for high-D data, all the high-D data are clustered into hexagons. In this step, the bin centroids ($C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2})$) (see @fig-meth Step 2) are obtained [@Carr2013].

@@

In this step, the algorithm proceeds to triangulate the hexagonal bin centroids (see @fig-meth Step 3). Triangulation is a fundamental process in computational geometry and computer graphics that involves dividing a set of points in a given space into interconnected triangles [@article30]. One common algorithm used for triangulation is Delaunay triangulation [@article26, @article54], where points are connected in a way that maximizes the minimum angles of the resulting triangles, leading to a more regular and well-conditioned triangulation.

@@

## Conclusion {#sec-conclusion}

<!--Our research introduces a comprehensive framework that leverages tours for interactive exploration of high-dimensional data coupled with a low-dimensional manifold, facilitated by the `quollr` R package. Regardless of the Non-Linear Dimension Reduction (NLDR) technique in use, our approach demonstrates effectiveness through simulation examples, particularly in the iterative removal of long edges for a smoother representation and capturing cluster variance.

@@

In conclusion, our framework presents a powerful tool for researchers and analysts in single-cell studies to assess their embeddings by visually inspecting them alongside the original data. By leveraging the advantages of tours and low-dimensional manifolds, our approach offers valuable insights into the data transformation process, empowering users to make informed decisions in analyzing high-dimensional data. Future work could enhance the method's robustness in the presence of noise and explore additional evaluation metrics for quantifying embedding accuracy.-->

### 2024-01-15 paper.tex
Rewrite the motivation of our paper

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt, sharp corners, frame hidden, breakable, enhanced]}{\end{tcolorbox}}\fi

@@

High-dimensional (high-D) data is widespread in many fields including
ecology and bioinformatics \citep{Guo2023}, in part because of new data
collection technologies \citep{Johnstone2009, ayesha2020overview}.
Working with high-dimensional data poses considerable challenges due to
the difficulty in visualizing beyond two dimensions \citep{Jia2022}.
High-dimensional data also presents difficulties for model fitting
\citep{Johnstone2009}, both computationally and interpretation, each of
which benefits from visualization.

To create visual representations of high-dimensional data, it is common
to apply dimension reduction techniques. Linear methods such as
principal component analysis (PCA) \citep{Karl1901} have been used for
many years, and remain popular. Non-linear methods such as
multi-dimensional scaling (MDS) \citep{Torgerson1967} have also been
routinely used. In the past decade, there has merged many new techniques
non-linear dimension reduction (NLDR), such as t-distributed stochastic
neighbor embedding (tSNE) \citep{Laurens2008}, uniform manifold
approximation and projection (UMAP) \citep{Leland2018}, designed to
capture the complex and non-linear relationships present within
high-dimensional data \citep{Johnstone2009}.

However, projecting high-dimensional data has limitations, such as
information loss and potential distortion of essential structures and
patterns \citep[\citet{article53}]{Jia2022}. The choice of technique and
parameters further impacts the accuracy of the visualization,
necessitating careful consideration for meaningful interpretation (see
Figure~\ref{fig-nldervis}).

Interactive and dynamic graphics systems have also been developed over
the years to enable visualizing high dimensions. One method, called a
tour \citep{Asimov1985}, shows a sequence of linear projections is shown
as a movie, allowing exploration without warping the space
\citep{lee2021review}. Interactive tools like \textbf{XGobi} and
\textbf{GGobi} have been successful in incorporating tours for exploring
high-dimensional data \citep{article60}. The R package \textbf{tourr}
\citep{article61} further enhances tour visualization within R, although
it may face limitations in frame rate and interactive features compared
to \textbf{GGobi}.

To overcome these limitations, the R package \textbf{detourr}
\citep{article22} has been developed, leveraging a Javascript widget via
htmlwidgets \citep{Ramnath2023} to achieve higher frame rates and
enhanced interactivity. Additionally, the R package \textbf{langevitour}
\citep{article09} utilizes Langevin Dynamics to generate a continuous
path of projections, eliminating the need for interpolation between
projections for animation. The tour technique has proven valuable in
exploring statistical model fits \citep{article58} and factorial
experimental designs \citep{article59}. Augmenting the results of
non-linear dimensional reduction methods with the tour, as demonstrated
in the \textbf{liminal} R package \citep{article21}, further enhances
data exploration.

While tours \citep{Asimov1985} preserve space without warping
\citep{lee2021review}, they require integrating multiple low-dimensional
views mentally to perceive high-dimensional structures. To address this
challenge, we propose a novel approach by combining the tour technique
with a low-dimensional manifold. This manifold is created through the
synergistic use of Non-Linear Dimension Reduction (NLDR) techniques,
hexagonal binning, and triangulation. By merging these techniques, our
approach offers a comprehensive and efficient means to visualize and
explore high-dimensional data while retaining the advantages of both
tours and NLDR. This integration facilitates a more intuitive perception
of complex data structures and empowers analysts with a robust tool for
assessing the accuracy of data transformations. The implementation of
our approach is available as an R package called \textbf{quollr}.

The outline of this paper is as follows. The
Section~\ref{sec-background} provides an detailed overview of dimension
reduction methods, triangulation, and tours. Building upon this
foundation, the Section~\ref{sec-methods} delves into the proposed
algorithm, \textbf{quollr}, and its implementation details. In
Section~\ref{sec-prediction}, discusses the effectiveness of the learned
low-dimensional manifold in accurately representing the complex
high-dimensional data. Following that, Section~\ref{sec-simpleex}
presents simple examples from simulations to illustrate the
functionality of the algorithm. Subsequently,
Section~\ref{sec-applications} showcases real-world applications of
\textbf{quollr} on different data sets, particularly in single-cell
RNA-seq data. These applications reveal insights into the performance
and trustworthiness of NLDR algorithms. We analyze the results to
identify situations where NLDR techniques may lead to misleading
interpretations. Finally, \textbf{?@sec-conclusions} concludes by
summarizing the findings and emphasizing the significance of the
proposed approach in tackling the challenges of high-dimensional data
visualization.

@@

As a result of hexagonal binning for high-D data, all the high-D data
are clustered into hexagons. In this step, the bin centroids
(\(C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2}\)) (see Figure~\ref{fig-meth}
Step 2) are obtained \citep{Carr2013}.

@@

In this step, the algorithm proceeds to triangulate the hexagonal bin
centroids (see Figure~\ref{fig-meth} (Step 3)). Triangulation is a
fundamental process in computational geometry and computer graphics that
involves dividing a set of points in a given space into interconnected
triangles \citep{article30}. One common algorithm used for triangulation
is Delaunay triangulation \citep[\citet{article54}]{article26}, where
points are connected in a way that maximizes the minimum angles of the
resulting triangles, leading to a more regular and well-conditioned
triangulation.

@@

\hypertarget{sec-discussion}{%
\section{Discussion}\label{sec-discussion}}

Our research introduces a comprehensive framework that leverages tours
for interactive exploration of high-dimensional data coupled with a
low-dimensional manifold, facilitated by the \texttt{quollr} R package.
Regardless of the Non-Linear Dimension Reduction (NLDR) technique in
use, our approach demonstrates effectiveness through simulation
examples, particularly in the iterative removal of long edges for a
smoother representation and capturing cluster variance.

In the example with doublets, our method successfully captures the tweak
within each cluster, indicating the variance present within them.
However, the model may not appear smooth in high-dimensional space due
to considerable noise when the data has a piecewise linear geometry,
such as the tree simulation.

The practical application of our framework, as showcased with the UMAP
view, enables visual inspection of well-separated clusters. Furthermore,
the combined tour and model provide a robust assessment of whether UMAP
preserves the data structure and accurately transforms the data.

The advantages of our approach include its versatility across various
NLDR techniques and the ability to generate interactive visualizations
for detailed exploration. The tour provides an intuitive way to navigate
and comprehend high-dimensional data while assessing the accuracy of
dimensionality reduction.

However, one limitation is that the approach may be less effective in
cases with significant noise, as seen in the tree simulation example.
Additionally, while our method aids in visual verification, quantifying
the accuracy of embeddings might require further evaluation metrics.

In conclusion, our framework presents a powerful tool for researchers
and analysts in single-cell studies to assess their embeddings by
visually inspecting them alongside the original data. By leveraging the
advantages of tours and low-dimensional manifolds, our approach offers
valuable insights into the data transformation process, empowering users
to make informed decisions in analyzing high-dimensional data. Future
work could enhance the method's robustness in the presence of noise and
explore additional evaluation metrics for quantifying embedding
accuracy.

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden, enhanced, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

High-dimensional (high-D) data is prevalent across various fields, such
as ecology and bioinformatics \citep{Guo2023}, due to advancements in
data collection technologies \citep{Johnstone2009, ayesha2020overview}.
However, visualization of high-D data introduces significant challenges,
because the complexity of visualizing data beyond two dimensions
\citep{Jia2022}. In recent years, interactive and dynamic graphics
systems like \textbf{liminal} \citep{article21} ---which employs
interactive tools like brushing and linking \citep{article58}---and
software tools such as \textbf{XGobi}, \textbf{GGobi} \citep{article60},
\textbf{tourr} \citep{article61}, \textbf{detourr} \citep{article22},
and \textbf{langevitour} \citep{article09}, involving dynamic methods
like tours \citep{Asimov1985}, have played a key role in visualizing
high-D data (data-vis).

To create low-dimensional representations (typically in 2D) of high-D
data, it is common to apply dimension reduction (DR) techniques.
Approaches for DR involve linear methods such as principal component
analysis (PCA) \citep{Karl1901}, non-linear methods such as
multi-dimensional scaling (MDS) \citep{Torgerson1967}. In the past
decade, many new non-linear dimension reduction (NLDR) techniques have
emerged, such as t-distributed stochastic neighbor embedding (tSNE)
\citep{Laurens2008} and uniform manifold approximation and projection
(UMAP) \citep{Leland2018}, designed to capture the complex and
non-linear relationships present within high-D data
\citep{Johnstone2009}. NLDR techniques are the 2D models of high-D data
in our context.

Visualization of NLDR techniques (m-vis) \citep{article59} for the same
high-D data is particularly important in understanding and finding the
best representation. If we have done so, the 2D models can be quite
different from the data structure shown in high-D space. Therefore,
visualizing the 2D model in high-D space (m-in-dis) is more useful to
answer different types of questions:

\begin{itemize}
\item
  Is there a best representation of high-D data or are they all
  providing equivalent information? Is there a best parameter choice to
  fit the 2D model? How does the model change when it's parameters
  change?
\item
  How well the does the 2D models capture the data structure? Is the
  model fitting able to capture different data structure like
  non-linear, clustering?
\end{itemize}

If we can't easily ask and answer these questions, the ability of
understanding the models are restricted. To find the best 2D model and
the parameter choices, better understanding of the underlying science is
really important.

Also, the importance of m-vis along with data-vis has been recognized
and incorporated into interactive software, \textbf{liminal}
\citep{article21}. But the 2D model and high-D visualize side by side
and interactive like brushing and linking connect the data in the two
panels. To address this challenge, we propose a novel approach by
combining the tour technique with a low-dimensional manifold. This
manifold is created through the synergistic use of NLDR techniques,
hexagonal binning, and triangulation. This integration facilitates a
more understanding of the data structure, how well (or how poorly) NLDR
techniques perform.

The outline of this paper is as follows. The
Section~\ref{sec-background} provides an detailed overview of dimension
reduction methods, and tours. Building upon this foundation, the
Section~\ref{sec-methods} delves into the proposed algorithm, its
implementation details, how to tune the model, model summaries, and a
synthetic example to illustrate the functionality of the algorithm.
Subsequently, Section~\ref{sec-applications} showcases applications of
the algorithm on different data sets, particularly in single-cell
RNA-seq data. These applications reveal insights into the performance
and trustworthiness of NLDR algorithms. We analyze the results to
identify situations where NLDR techniques may lead to misleading
interpretations. Finally, \textbf{?@sec-conclusions} concludes by
summarizing the findings and emphasizing the significance of the
proposed approach in tackling the challenges of high-dimensional data
visualization.

@@

As a result of hexagonal binning for high-D data, all the high-D data
are clustered into hexagons. In this step, the bin centroids
(\(C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2})\)) (see Figure~\ref{fig-meth}
Step 2) are obtained \citep{Carr2013}.

@@

In this step, the algorithm proceeds to triangulate the hexagonal bin
centroids (see Figure~\ref{fig-meth} Step 3). Triangulation is a
fundamental process in computational geometry and computer graphics that
involves dividing a set of points in a given space into interconnected
triangles \citep{article30}. One common algorithm used for triangulation
is Delaunay triangulation \citep[\citet{article54}]{article26}, where
points are connected in a way that maximizes the minimum angles of the
resulting triangles, leading to a more regular and well-conditioned
triangulation.

@@

\hypertarget{sec-conclusion}{%
\section{Conclusion}\label{sec-conclusion}}
